Semi-quantitative petrographic characterization of Bronze Age quartzite tools – a multivariate approach

### 2021-10-19 analysis/paper/paper.Rmd
Rewritten text, added some visualisation to figures

---
title: "Semi-quantitative petrographic characterization of Bronze Age quartzite tools – a multivariate approach"
author:
  - Mattias Sjölander:
      email: mattias.sjolander@umu.se
      institute: [UmU]
      correspondence: true
institute:
  - UmU: Umeå University
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
mainfont: Times New Roman
bibliography: references.bib
csl: "../templates/journal-of-archaeological-and-anthropological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  Spectroscopy; Quartzite; Characterization; Lithics
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

@@

```{r load-packages-and-data}
library(ChemoSpec)
library(data.table)
library(factoextra)
library(ggbiplot)
library(ggplot2)
library(ggsci)
library(gridExtra)
library(missMDA)
library(patchwork)
library(prospectr)
library(tidyverse)
library(spectrolab)

@@

#Filter xrf, nir and raman data to focus on points and preforms made from quartz/quartzite material 
Points.xrf <-
  XRF.csv %>%
    filter(material == "Quartz" | material == "Quartzite" | material == "Breccie quartz" | material == "Breccie quartzite" | material  == "Felsitic porphyric quartz",
         type == "Point" | type == "Point fragment" | type == "Preform")

Points.nir <-
  nir.csv %>%
    filter(material == "Quartz" | material == "Quartzite" | material == "Breccie quartz" | material == "Breccie quartzite" | material  == "Felsitic porphyric quartz",
         type == "Point" | type == "Point fragment" | type == "Preform")

Points.raman <-
  raman.csv %>%
    filter(material == "Quartz" | material == "Quartzite" | material == "Breccie quartz" | material == "Breccie quartzite" | material  == "Felsitic porphyric quartz", type == "Point" | type == "Point fragment" | type == "Preform")

#Fill the NA fields in the munsell_hue column and mark them as translucent samples
Points.nir[8][is.na(Points.nir[8])] <- "Translucent"
Points.raman[8][is.na(Points.raman[8])] <- "Translucent"
Points.xrf[20][is.na(Points.xrf[20])] <- "Translucent"

#prepare color/fill and symbols for score plots
pca.colors <- c("#E31CFF", "#718EFF", "#0080FF", "#D5DF9E", "#C6DF39", "#FFFFC1", "#FFFFA1", "#FFFF00", "#FFFF38", "#FFB468", "#000000", "#202020", "#404040", "#606060", "#808080", "#9F9F9F", "#DFDFDF", "#FFFFFF")
pca.shapes <- c(24,25,23,22,21)
pca.munsell_hue <- c("Translucent", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9")

```
<!-- The actual document text starts here: -->

@@

In regions where local flint and obsidian sources are non-existent, such as Northern Fennoscandia, other materials like quartz, quartzite and slate have been the primary target for lithic production^[This is not to say that flint was not available or used in this region] [@baudou78; @rankama_etal06; @olausson_etal12]. Non-flint lithics and raw material studies in  European archaeology have seen a steady increase during the last decades, providing much needed development in a previously neglected field [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; -@prieto_etal20; -@prieto_etal21; @ramacciotti_etal19]. With the lack of any natural geological source of flint in Northern Sweden both quartz and quartzite are instrumental for the inference of settlement and mobility within the region. While quartz and quartzite technology have been the target of a number of studies over the years, there have been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10].

One reason for the lack of methodological development could be the complex geological background of the region and the abundance of moraine formations [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16]. The degree to which geological materials have been mixed and transported demands considerable effort in order to reconstruct the events that have formed the modern day landscape. While tentaive interpretations of lithic material sourcing tends to describe strategies leaning towards direct procurement from the mountain zone [@forsberg85; -@forsberg12], both quartz and quartzite could potentially have been sourced locally from exposed veins in the bedrock, or as loose rocks and pebbles in the numerous moraine formations as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06]. Due to this ambigous nature of availability of raw materials lprovenience studies are increasingly complex and fraught with uncertainty.

The defintition for what constitutes a *quartzite* is not so stratightforward. Prieto et al. [-@prieto_etal19 p.15; -@prieto_etal20 p.32] notes that the usage of the term *quartzite* in archaeology has been somewhat ambiguous, often used interchangeably between sedimentary orthoquartzite and metamorphic quartzite. The defintion of quartzite has been a topic of discussion in geology as well [@howard05], and the definition hence varies somewhat between authors. Visual characterization of quartzite on a megascopic level is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures, but Howard also emphasises the need for field-based methodologies.

In recent years there has been increased development of petrographic characterization of quartzite using a combination of destructive methods and spectroscopy [e.g. @blomme_etal12; @cnudde_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20]. These have had success in distinguishing between different groups of quartzite types, allowing for more meaningful studies into quartzite-based resource management strategies. In archaeology there are times, however, when destructive techniques simply are not applicable. Artefact collections (museum, private, etc.) represent a valuable resource, as well as one of the easier ways of gaining access to large datasets with some geographical distribution. While destructive methods may prohibit access to such material there are still non-destructive methods that can generate valuable information. Aside from being non-destructive spectroscopic instrumentation feature fast measurements, making it an attractive alternative for field-based screening purposes. While X-Ray Fluorescence (XRF) is the go-to spectroscopic technique for petrographic characterization [@shackley11; @tykot03], due to its capability of providing compositional data of the elemental content of a material, there are other instrumentations that may provide a deeper insight into the structural properties. Near Infrared spectroscopy (NIRS) goes almost as far back as the XRF [@hunt77], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@linderholm_geladi14a; @linderholm_geladi14b; @sciuto19]. The same can be said for Raman spectroscopy, which has been applied extensively in the Cultural Heritage sector and the analysis of artwork and manuscripts [@vandenabeele12].

In the current study a dataset of quartzite artefacts from museum collections in Västerbotten, Sweden, have been analysed using the three above mentioned spectroscopic instrumentations (NIRS, Raman, XRF). They all provide different information on the structure and composition of the material and will thus be evaluated in regards to the different features they record in the quartz/quartzite. This will be achieved via Principal Component Analysis (PCA) where the observed groupings of the material will be related to the spectral features. Lastly, the potential for combining the results of the analyses in order to study joint relations will be discussed.

# Material
```{r select-point-data}
#Filter data to focus on points, point fragments and preforms made from quartz/quartzite materials
Points2 <-
  samples %>%
    filter(material == "Quartz" | material == "Quartzite" | material == "Breccie quartz" | material == "Breccie quartzite" | material  == "Felsitic porphyric quartz",
         type == "Point" | type == "Point fragment" | type == "Preform")

@@

#Then filter to focus on quartz/quartzite material that are points and preforms 
lycksele_301 <-
  lycksele %>% 
    filter(material == "Quartz" | material == "Quartzite" | material == "Breccie quartz" | material == "Breccie quartzite" | 
             material  == "Felsitic porphyric quartz", type == "Point" | type == "Point fragment" | type == "Preform")

@@

The current study is based on spectroscopic surface measurements on quartzite points found within the County of Västerbotten, Sweden. A dataset of `r length(sample_uniq)` points, point fragments and preforms made from quartz/quartzite material stored at the collections at Västerbotten and Skellefteå museums were analysed using three different spectroscopic instrumentations (Appendix XX). The colour of the material has a mainly black (N1-N3), grey or white (N9) hue, as well as opaque and almost translucent material with a beige tint. The colour of the material has been classified both according to the munsell system, as well as sorted into general categories of hue (*dark* ~ N1-N4, *light* ~ N5-N8, *translucent*, *white* ~ N9). A number of these points have, according to prior classifications, been worked using bifacial technology. While there currently is no local chronology available for this artefact type in the North Swedish region, previous research suggest they were in use during the 4 000 – 2 000 BP period [@forsberg85; @forsberg89; @mjaerum12]. Similarly, early studies of raw material use at sites dating to this period seems to indicate an increase in the use of quartzite compared to previous periods [@baudou78].

A large portion of the collections stored at North Swedish museums were recovered from excavations and surveys performed in conjunction with the river regulations during the 1940s – 1980s [@biornstad06; @janson_hvarfner60], as well as subsequent follow-up survey projects [e.g. @andersson15]. This has likely had a contributing factor to the distribution pattern of the artefacts observed in fig. \@ref(fig:distribution-map) as the known sites with finds of points and preforms are concentrated around the parishes of Dorotea, Lycksele, Vilhelmina and Tärnaby. This can be seen more clearly in fig. \@ref(fig:distribution-by-parish) with the four parishes accounting for almost ¾ of the total no. of samples.

Västerbotten museum have come far in their digitization effort, making large portions of their collections searchable. Based on a query of all points labelled as *quartz*, *brecciated quartz* or *quartzite* a list was generated from their database. Sampling of this pool of artefacts was not random, and based on geographical distribution, in an effort to include as many artefacts as possible from areas with few finds available, such as along the Vindel river. None of these were from the northernmost Skellefte river, however, and thus a selection of 5 artefacts each from three well-known sites stored at the collections of Skellefte museum were included. This resulted in a total dataset of `r length(tot_uniq)` artefacts, and after documentation and looking up missing information, `r length(sample_uniq)` of these were determined to be either points or preforms made from quartz/quartzite material.

@@

While parts of the collections at Västerbotten museum have been petrographically analysed by a geologist [@biornstad68; @ahman67], at this point it is difficult to say with any certainty how many (if any) of the artefacts in the current dataset were among these. It is fair to assume that the vast majority have been classified visually based on megascopic features according to a “standard” archaeological nomenclature. In the dataset the four main types of registered materials are: quartz, quartzite, brecciated quartz^[Original Swedish terms are *brecciekvarts*/*brecciekvartsit* , the latter possibly a spelling mistake], and brecciated quartzite. Those that had no prior material classification registered were sorted into either the quartz or quartzite category. The reason for including ‘quartz’ in the study is in order to test whether it is possible to separate quartz from quartzite using these methods, as well as a way to account for potential mistakes in the original classification.

@@

\linebreak
 \linebreak
All measurements were performed at the Environmental Archaeology Lab at Umeå University using three different instrumentations; NIRS, Raman scattering, and XRF. The analysis was conducted in a dark room in order to minimize light pollution, with the main light sources in the room being the computer screens. Each artefact was measured twice on different sides, with additional measurements in cases where sampling was difficult. The only exception to this was the XRF, where due to time constraints one measurement was collected as a rule apart from a few exceptions. This lowers the statistical reliability of the XRF analysis and inferences of the results should be evaluated with this in mind. The size of the dataset alleviates some of these issues, as for the larger sites there are multiple objects that may be of the same material, however, for the smaller sites there still remains some uncertainties.

Pre-processing of the spectra is different for each instrumentation and based on relevant literature. After pre-processing the data was modelled using PCA, which is an unsupervised exploratory technique for reducing the dimensionality of the dataset (Geladi and Linderholm, 2020). PCA modelling was performed using the base R *prcomp()* method [@Rteam17].

@@

NIR analysis was conducted with the Analytical Spectral Device (ASD) LabSpec 4, using a contact probe. The ASD features a detection range of 350 – 2500 nm, with spectral sampling (resp. spectral resolution) of 1.4 nm (resp. 3 nm) in the visible and near infrared range and 1.1 nm (resp. 10 nm) in the short‐wave infrared range. Each object was measured against a white reference background and the collected measurements were averaged for each artefact. Two parts of the NIR spectra were analysed separately, the visual region (400 – 750 nm) and the 1 000 – 2 500 nm region. In both analyses the data was mean-centred before PCA modelling.

@@

The Raman instrumentation used was a portable i-Raman EX featuring a 1064 nm excitation laser. The spectral coverage of the instrumentation is 175 – 2 500 cm-1 ^[Raman shifts are typically given in wavenumber (cm-1), instead of spectral wavelength (nm)] with a resolution of 9.5 cm-1. Access to the spectrometer was made possible by the department of Forest Biomaterials and Technology at the Swedish University of Agricultural Sciences (SLU) in Umeå.

@@

## X-Ray Fluourescence
XRF measurements were collected using a Thermo Scientific Niton XL5 Analyzer, connected to a Thermo ScientificTM portable test stand. The reference calibration used for element quantification was set to *mining* mode. Sampling with this setting takes 2 minutes in order for the instrument to cycle through all of the filters (main, low, high, light). Due to time constraints and the number of objects in the current dataset it was only possible to collect one measurement on each object, apart from some exceptions. For materials that featured high heterogeneity the sample point was focused, as much as possible, on the surrounding matrix. The XRF data was log-transformed, mean-centered and scaled to unit variance before PCA modelling. As the resulting dataset also features values below the Limit of Detection (LOD), these were treated as NULL, or missing values, during PCA. Since missing values are not permitted during PCA these had to imputed before the analysis, which was performed using the *missMDA* package [@josse_husson16] developed for datasets with incomplete data. Two methods for imputation can be implemented, iterative or regularized, since the former suffers when there are a large number of missing values the latter was chosen.

@@

<!--```{r vis-nir-pca, fig.width=18, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of 1st and 2nd components for the 400 nm – 750 nm region of the spectrum. Quartzite with a darker surface group to the right in the model, while white objects group in the bottom left quadrant."}
knitr::include_graphics('./analysis/figures/fig_4_1_visnir_averaged_pca_p1_p2.jpg')
```-->

@@

#bind the basic fviz plot for PC 1 and 2, and use as basis for a more customizeable plot in ggpplot
ggplot(cbind(basic_plot1$data, Points.nir[, c(8,10)]),
             aes(x=x, y=y, 
                 shape = material, 
                 fill = munsell_hue)) +
  geom_point(size=3) +
  theme_bw() +
  ggtitle("Near infrared 400 - 750 nm") +
  labs(x = vis.pc1lab,
       y = vis.pc2lab) +
  scale_shape_manual(name = "Material",
                     values=pca.shapes) +
  scale_fill_manual(name = "Munsell hue",
                    breaks = c("Translucent", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
                    values=pca.colors) +
  guides(shape = guide_legend(override.aes = list(fill = "black"), 
                              ncol=2, 
                              title.position="top", 
                              title.hjust = 0.5),
         fill = guide_legend(override.aes = list(shape = 22,
                                                 fill = pca.colors,
                                                 color = "black",
                                                 size=3),
                             title.position="top", 
                             title.hjust = 0.5)) +
  theme(plot.title = element_text(size = 12, face = "bold", colour = "black", vjust = - 10, hjust = 0.02),
        axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
        axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
        legend.title = element_text(size = 12, face = "bold", colour = "black"),
        legend.background = element_rect(linetype = "solid", color = "black"),
        legend.position = "bottom")
```

@@

vis.translucent <- Points.nir %>% 
  filter(hue == "Translucent") %>% 
  select(c(1:28),c(79:429))

@@

plot(mean(spec.vd), lwd = 1.2, col = "black", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Dark")

@@

plot(mean(spec.vt), lwd = 1.2, col = "purple", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Translucent")

plot(mean(spec.vw), lwd = 1.2, col = "blue", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - White")
```

<!--```{r vis-nir-spectra, fig.width=15, fig.height=5, fig.fullwidth=TRUE, fig.cap="Average spectra for dark (N1), translucent (MV) and white (N1) groups in the 400 - 750 nm region. Dark material has a higher overall absorbance, whereas white and translucent material feature lower absorbance. Brecciated quartz, which is typically described as a translucent material, falls somewhere inbetween the above groups."}
knitr::include_graphics('./analysis/figures/fig_4_2_breccie_averaged_visnir_spectra.jpg')
```-->

@@

## NIR - near infrared range
No distinct peaks occurred in the 600 – 1 000 nm band, and thus the analysis focused on the 1 000 – 2 500 nm band to lessen the impact of noise. In the case of NIRS similar features can be observed for all four colour groups, with darker materials exhibiting higher absorbance and less defined peaks. The presence of molecular water can be seen in the peaks at 1 400 nm and 1 900 nm  [@hunt77; @sciuto_etal19] (fig. \@ref(fig:nir-spectra)). Peaks also occur at 2 200 nm and 2 350 nm, where the former has been related to combination vibrations of Al-OH and the latter falls within one of the stronger CO^3^ bands [@clark_etal90]. Inspection of individual spectra for the *light* group, however, also reveals a 5th peak for some samples at 2 450 nm. Comparisons to spectra collected on muscovite ([@clark_etal90] p.12 675) shows similar peaks located at 2 200 nm, 2 350 nm, and 2 450 nm. It is therefore possibly that the 2 350 nm band in some cases corresponds to CO^3^ vibrations and in other cases Al-OH.

>>>>

---
title: "Semi-quantitative petrographic characterization of Bronze Age quartzite tools – a multivariate approach"
author:
  - Mattias Sjölander:
      email: mattias.sjolander@umu.se
      institute: [UmU]
      correspondence: true
institute:
  - UmU: Umeå University
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" 
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
mainfont: Times New Roman
bibliography: references.bib
csl: "../templates/journal-of-archaeological-and-anthropological-science.csl"
abstract: |
  Text of abstract
keywords: |
  Spectroscopy; Quartzite; Characterization; Lithics
highlights: |
  These are the highlights. 
---

@@

<!-- The following code chunk defines prepares most of the data that will be needed and reads in the necessary libraries -->

```{r load-packages-and-data}
library(ChemoSpec)
library(data.table)
library(factoextra)
library(ggbiplot)
library(ggplot2)
library(ggsci)
library(grid)
library(gridBase)
library(gridExtra)
library(missMDA)
library(patchwork)
library(prospectr)
library(tidyverse)
library(spectrolab)

@@

#Filter xrf, nir and raman data to focus on points and preforms made from quartz/quartzite material 
Points.xrf <-
  XRF.csv %>%
    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
         type == "Point" | type == "Point fragment" | type == "Preform")

Points.nir <-
  nir.csv %>%
    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
         type == "Point" | type == "Point fragment" | type == "Preform")

Points.raman <-
  raman.csv %>%
    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz", type == "Point" | type == "Point fragment" | type == "Preform")

#Fill the NA fields in the munsell_hue column and mark them as translucent samples
Points.nir[8][is.na(Points.nir[8])] <- "Colourless"
Points.raman[8][is.na(Points.raman[8])] <- "Colourless"
Points.xrf[20][is.na(Points.xrf[20])] <- "Colourless"

#prepare color/fill and symbols for score plots
pca.colors <- c("#E31CFF", "#718EFF", "#0080FF", "#D5DF9E", "#C6DF39", "#FFFFC1", "#FFFFA1", "#FFFF00", "#FFFF38", "#FFB468", "#000000", "#202020", "#404040", "#606060", "#808080", "#9F9F9F", "#DFDFDF", "#FFFFFF")
pca.shapes <- c(24,25,23,22,21)
pca.munsell_hue <- c("Colourless", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9")

@@

In regions where local flint and obsidian sources are non-existent, such as Northern Fennoscandia, other materials have been the primary target for lithic production^[This is not to say that flint was not available or used in these regions] [@baudou78; @rankama_etal06; @olausson_etal12]. The last decades have seen an increase in interest for non-flint lithics and raw material studies in  European archaeology, providing much needed development in a previously neglected field [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19]. With the lack of any natural geological source of flint in Northern Sweden both quartz and quartzite are instrumental for the inference of settlement and mobility within the region. While quartz and quartzite technology have been part of a number of studies over the years, there have been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10]. One reason for the lack of methodological development in North Swedish research could be the complex geological background of the region and the abundance of moraine formations [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16]. The degree to which geological materials have been mixed and transported demands considerable effort in order to reconstruct the events that have formed the modern day landscape. While tentaive interpretations of lithic material sourcing tends to describe strategies leaning towards direct procurement from the mountain zone [@forsberg85; -@forsberg12], both quartz and quartzite could potentially have been sourced locally from exposed veins in the bedrock, or as loose rocks and pebbles in the numerous moraine formations, as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06 pp.249-250]. Due to this ambiguous nature of availability of raw materials provenience studies are complex and fraught with uncertainty.

Attempting to define what constitutes a *quartzite* is not so stratightforward. Prieto et al. [-@prieto_etal19 p.15; -@prieto_etal20 p.32] notes that the usage of the term *quartzite* in archaeology has been somewhat ambiguous, often used interchangeably between sedimentary orthoquartzite and metamorphic quartzite. The defintion of quartzite has been a topic of discussion in geology as well [@howard05], and it hence varies somewhat between authors. Visual characterization on a megascopic level is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures, but Howard also emphasises the need for field-based methodologies. There are also situations where destructive techniques simply are not applicable. Artefact collections (museum, private, etc.) represent a valuable resource, as well as one of the easier ways of gaining access to large datasets with some geographical distribution. While destructive methods may prohibit access to such material there are still non-destructive methods that can generate valuable information. Aside from being non-destructive spectroscopic instrumentation feature fast measurements, making it an attractive alternative for field-based screening purposes. While X-Ray Fluorescence (XRF) is the go-to spectroscopic technique for petrographic characterization [@shackley11; @tykot03], due to its capability of providing compositional data, there are other instrumentations that may provide a deeper insight into the structural properties. The use of near-infrared spectroscopy (NIRS) goes almost as far back as the XRF [@hunt77], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@linderholm_geladi14a; @linderholm_geladi14b; @sciuto19]. The same can be said for Raman spectroscopy, which has been applied extensively in the Cultural Heritage sector and in the analysis of artwork and manuscripts [@vandenabeele12].

In the current study a dataset of quartzite artefacts from museum collections in Västerbotten, Sweden, have been analysed using the three above mentioned spectroscopic instrumentations (NIRS, Raman, XRF). The goal is to evaluate their capabilities for fingerprinting quartz and quartzite in the North Swedish region. This will be achieved via Principal Component Analysis (PCA), where the observed groupings of the material in the model will be related to the spectral features for each instrumentation. 
 \linebreak

# Material
```{r select-point-data}
#Filter data to focus on points, point fragments and preforms made from quartz/quartzite materials
Points2 <-
  samples %>%
    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
         type == "Point" | type == "Point fragment" | type == "Preform")

@@

#Then filter to focus on quartz/quartzite material that are points and preforms 
lycksele_301 <-
  lycksele %>% 
    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | 
             material  == "Felsitic porphyric quartz", type == "Point" | type == "Point fragment" | type == "Preform")

@@

The current study is based on spectroscopic surface measurements on quartz and quartzite points recovered from sites within the County of Västerbotten, Sweden. A dataset of `r length(sample_uniq)` points, point fragments and preforms made from quartz/quartzite material stored at the collections at Västerbotten (www.vbm.se) and Skellefteå (www.skellefteamuseum.se) museums were analysed using NIRS, raman and XRF spectroscopy (Appendix XX). The material features a mainly black (N1-N3), grey or white (N9) hue, as well as a subset of semi-translucent material with a beige tint to almost fully translucent. The colour of the material has been classified both according to the munsell system, as well as sorted into general categories of hue (*dark* ~ N1-N4, *light* ~ N5-N8, *white* ~ N9, *Colourless* ~ lack of colour, opaque to translucent) for visualization purposes. A number of these points have, according to prior classifications, been worked using bifacial technology. While there currently is no local chronology available for this artefact type in the North Swedish region, previous research suggest they were in use during the 4 000 – 2 000 BP period [@forsberg85; @forsberg89; @mjaerum12]. Similarly, early studies of raw material use at sites dating to this period seems to indicate an increase in the use of quartzite compared to previous periods [@baudou78].

A large portion of the collections stored at North Swedish museums were recovered from excavations and surveys performed in conjunction with the river regulations during the 1940s – 1980s [@biornstad06; @janson_hvarfner60], as well as subsequent follow-up survey projects [e.g. @andersson15]. This has likely had a contributing factor to the distribution pattern of the artefacts observed in fig. \@ref(fig:distribution-map) as the known sites with finds of points and preforms are concentrated around the parishes of Dorotea, Lycksele^[In this case the majority of the samples from Lycksele are from the rock painting site of Lycksele 301], Vilhelmina and Tärnaby. This can be seen more clearly in fig. \@ref(fig:distribution-by-parish) with the four parishes accounting for almost ¾ of the total no. of samples.

Västerbotten museum have come far in their digitization effort, making large portions of their collections searchable online. Based on a query of all points labelled as *quartz*, *brecciated quartz* or *quartzite* a list was generated from their database. Sampling of this pool of artefacts was not random, and based on geographical distribution, in an effort to include as many artefacts as possible from areas with few finds available, such as along the Vindel river. None of these were from the northernmost Skellefte river, however, and thus a selection of 5 artefacts each from three well-known sites stored at the collections of Skellefte museum were included. This resulted in a total dataset of `r length(tot_uniq)` artefacts, and after documentation and looking up missing information, `r length(sample_uniq)` of these were determined to be either points or preforms made from quartz/quartzite material.

@@

While parts of the collections at Västerbotten museum have been petrographically analysed by a geologist [@biornstad68; @ahman67], at this point it is difficult to say with any certainty how many (if any) of the artefacts in the current dataset were among these. It is fair to assume that the vast majority have been classified visually based on megascopic features according to a “standard” archaeological nomenclature. In the dataset the four main types of registered materials are: quartz, quartzite, brecciated quartz^[Original Swedish terms are *brecciekvarts*/*brecciekvartsit* , the latter possibly a spelling mistake], and brecciated quartzite. Those that had no prior material classification registered were sorted into either the quartz or quartzite category. The reason for including *quartz* in the study is in order to test whether it is possible to separate quartz from quartzite using these methods, as well as a way to account for potential mistakes in the original classification.

@@

\linebreak
 \linebreak
All measurements were performed at the Environmental Archaeology Lab at Umeå University. The analysis was conducted in a dark room in order to minimize light pollution, with the main light sources in the room being the computer screens. Each artefact was measured twice on different sides, with additional measurements in cases where sampling was difficult. The only exception to this was the XRF, where due to time constraints one measurement was collected as a rule apart from a few exceptions. This lowers the statistical reliability of the XRF analysis and inferences of the results should be evaluated with this in mind. The size of the dataset alleviates some of these issues, as for the larger sites there are multiple objects that may be of the same material, however, for the smaller sites there still remains some uncertainties. For materials that featured high heterogeneity the sample point was focused, as much as possible, on the surrounding matrix.

Pre-processing of the spectra is different for each instrumentation and based on relevant literature. After pre-processing the data was modelled using PCA, which is an unsupervised exploratory technique for reducing the dimensionality of the dataset [@geladi_linderholm20]. PCA modelling was performed using the base R *prcomp* function [@Rteam17; @rcoreteam21].

@@

NIR analysis was conducted with the Analytical Spectral Device (ASD) LabSpec 4, using a contact probe. The ASD features a detection range of 350 nm – 2500 nm, with spectral sampling (resp. spectral resolution) of 1.4 nm (resp. 3 nm) in the visible and near infrared range and 1.1 nm (resp. 10 nm) in the short‐wave infrared range. Each object was measured against a white reference background and the collected measurements were averaged for each artefact. Two parts of the As there were no notable features in the visible range the analysis was restricted to the 1 000 nm – 2 500 nm region in order to lessen the impact of noise. The data was mean-centred before PCA modelling.

@@

The Raman instrumentation used was a portable i-Raman EX featuring a 1064 nm excitation laser. The spectral coverage of the instrumentation is 175 cm-1 – 2 500 cm-1 ^[Raman shifts are typically given in wavenumber (cm-1), instead of spectral wavelength (nm)] with a resolution of 9.5 cm-1. Access to the spectrometer was made possible by the department of Forest Biomaterials and Technology at the Swedish University of Agricultural Sciences (SLU) in Umeå.

@@

## X-Ray Fluourescence
XRF measurements were collected using a Thermo Scientific Niton XL5 Analyzer, connected to a portable test stand. The reference calibration used for element quantification was set to *mining* mode. Sampling with this setting takes 2 minutes in order for the instrument to cycle through all of the filters (main, low, high, light). Due to time constraints and the number of objects in the current dataset it was only possible to collect one measurement on each object, apart from some exceptions. The XRF data was log-transformed, mean-centered and scaled to unit variance before PCA modelling. As the resulting dataset also features values below the Limit of Detection (LOD), these were treated as NULL, or missing values, during PCA. Since missing values are not permitted during PCA these had to imputed before the analysis, which was performed using the *missMDA* R package [@josse_husson16] developed for datasets with incomplete data. Two methods for imputation can be implemented, iterative or regularized, since the former suffers when there are a large number of missing values the latter was used.

@@

@@

#bind the basic fviz plot for PC 1 and 2, and use as basis for a more customizeable plot in ggpplot
ggplot(cbind(basic_plot1$data, Points.nir[, c(8,10)]),
             aes(x=x, y=y, 
                 shape = material, 
                 fill = munsell_hue)) +
  geom_point(size=3) +
  theme_bw() +
  ggtitle("Near infrared 400 - 750 nm") +
  labs(x = vis.pc1lab,
       y = vis.pc2lab) +
  scale_shape_manual(name = "Material",
                     values=pca.shapes) +
  scale_fill_manual(name = "Munsell hue",
                    breaks = c("Colourless", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
                    values=pca.colors) +
  guides(shape = guide_legend(override.aes = list(fill = "black"), 
                              ncol=2, 
                              title.position="top", 
                              title.hjust = 0.5),
         fill = guide_legend(override.aes = list(shape = 22,
                                                 fill = pca.colors,
                                                 color = "black",
                                                 size=3),
                             title.position="top", 
                             title.hjust = 0.5)) +
  theme(plot.title = element_text(size = 12, face = "bold", colour = "black", vjust = - 10, hjust = 0.02),
        axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
        axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
        legend.title = element_text(size = 12, face = "bold", colour = "black"),
        legend.background = element_rect(linetype = "solid", color = "black"),
        legend.position = "bottom")
```

@@

vis.translucent <- Points.nir %>% 
  filter(hue == "Colourless") %>% 
  select(c(1:28),c(79:429))

@@

plot(mean(spec.vd), lwd = 1.2, col = "black", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Dark"); vps1 <- do.call(vpStack, baseViewports())

@@

plot(mean(spec.vt), lwd = 1.2, col = "purple", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Colourless")

plot(mean(spec.vw), lwd = 1.2, col = "blue", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - White"); vps4 <- do.call(vpStack, baseViewports())
```

@@

## NIR - near infrared range
In the case of NIRS similar features can be observed in the average spectra for all four colour groups, with darker materials exhibiting higher absorbance and the light group the most distinct peaks. The presence of molecular water can be seen in the peaks at 1 400 nm and 1 900 nm  [@hunt77; @sciuto_etal19] (fig. \@ref(fig:nir-spectra)). Peaks also occur at 2 200 nm and 2 350 nm, where the former has been related to combination vibrations of Al-OH and the latter falls within one of the stronger CO^3^ bands [@clark_etal90]. Inspection of individual spectra for the *light* group, however, also reveals a 5th peak for some samples at 2 450 nm. Comparisons to spectra collected on muscovite ([@clark_etal90] p.12 675) shows similar peaks located at 2 200 nm, 2 350 nm, and 2 450 nm. It is therefore possible that the 2 350 nm band in some cases corresponds to CO^3^ vibrations and in other samples Al-OH.

The PCA shows a strong trend along PC 1 that seems to correlate with the colour of the material(fig. \@ref(fig:nir-pca)). Darker quartzite cluster to the right with white and translucent samples grouping to the opposite side. While the white samples seem to gravitate to the bottom left of the model the samples that are somewhat translucent, such as the *brecciated quartz*, seem to cluster towards the top-left. A small set of 3 black samples at the top of the model separate themselves clearly from the rest of the data points. These three samples come from different sites and exhibit no peaks in their spectra, making them an interesting set of outliers.
